Investigating soundscapes perception through acoustic scenes simulation.

### 2017-08-23 lafaySimulatedScenePerception.tex
on track

\newcommand{\gl}[1]{\textcolor{blue}{GL : #1}}
\newcommand{\nm}[1]{\textcolor{magenta}{#1}}
\newcommand{\ml}[1]{\textcolor{red}{#1}}
\newcommand{\jlc}[1]{\textcolor{green}{#1}}
\newcommand{\ie}{\emph{i.\,e.}}
\newcommand{\Ie}{\emph{I.\,e.}}
\newcommand{\cf}{cf.}
\newcommand{\Cf}{Cf.}
\newcommand{\vs}{\emph{vs.}}
\newcommand{\Vs}{\emph{Vs.}}
\newcommand{\al}{\emph{et~al.}}
\newcommand{\eg}{\emph{e.\,g.}}
\newcommand{\Eg}{\emph{E.\,g.}}
\newcommand{\etc}{\emph{etc.}}
\newcommand{\myfloatalign}{\centering}

@@

    \author{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J.F. Petiot$^{1)}$}

@@

First, a subject can be asked to describe a given sound following a given perceptual scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}. The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}. That being said, those stimuli respectively are or can be recorded and analysed. It thus allows the experimenter to gather a coarse grain physical description of the stimuli.

@@

Second, a subject can be asked to describe a given sound environment \cite{guastavino2006ideal, dubois2006cognitive}. A large amount of quantitative and semantic information is then collected about the representation of this type of sound environment that the subject has. Unfortunately, without any reference to sound data, this representation can hardly be characterized physically.

@@

For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sounds sources that compose a sound environment of interest. With such a material, not only the type of sound sources is indeed available for study, but also the exact level and audio waveform for each source together with the structural characteristics of the scene, that is the temporal distribution of the events.

@@

The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}. The experiments  1.a \emph{simulation} and  1.b \emph{pleasantness evaluation} are presented in Section \ref{sec:simulation}. Experiment 2 \emph{pleasantness evaluation after modification of the scenes} is presented in Section \ref{sec:modification}. Conclusions and discussion about future work follow in Section \ref{sec:conclusion}. Experiment 1.a is based on a pilot study described in \cite{lafay2014new}.

@@

\emph{Simscene} is an online digital audio tool whose first version have been developed as part of the HOULE project\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}. It has been designed to run on the popular web browsers \emph{Chrome} and \emph{Firefox}. It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digita audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}. The interface for selecting the sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.

@@

\emph{Simscene} is designed as a simplified audio sequencer, with sequencing parameters specifically chosen for the generation of realistic soundscapes. To do so, the user first selects a sound source using a non verbal selection interface presented in Section~\ref{sec:simscene_ssf}. A track is then created for this sound source within the simulator interface. The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sounds source. Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) to provide some comments about the simulated scene.

@@

The literature usually distinguish between:
\begin{itemize}
\item {a sound event}: an isolated sound, sparsely distributed whose acoustical characteristics may change with respect to time
\item {a sound texture}: an isolated sound of long duration whose acoustical characteristics are stable with respect to time \cite{saint1995classification}.
\end{itemize}

The simulation tool \emph{Simscene} follows this distinction and consider two distinct sound databases: one with the classes of events only and the other with classes of textures only. These two types also have specific scheduling procedures during the simulation process.

@@

Maffiolo \cite{maffiolo_caracterisation_1999} distinguish two distinct categorization processes, that differentiate themselves with the capacity of the listener to identify sound events. She shows that those processes leads to two abstract cognitive category respectively termed "the event sequences" and "the amorphous sequences".

@@

The event sequences are composed of salient events that can easily be recognized such as \emph{car start} or \emph{male speech}. They arise from a descriptive analysis based on the identification of the sound sources. On contrary, the amorphous sequences are sound environments where distinct events can hardly be found, such as \emph{traffic hubhub}. They result from a holistic analysis based on global acoustic features.

@@

\begin{figure}[t]
        \myfloatalign
        \includegraphics[width=.8\linewidth]{gfx/ch_4/3.png}
       \caption{Hierachical organisation of the isolated sounds used of the simulation.}\label{fig:orgDb}
\end{figure}

@@

@@

\begin{itemize}
\item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni- scenes is also present in terms of perceived pleasantness, we look at the existence of differences between $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$ with respect to the two types of scenes. The significativity is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$;
\item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni- scenes has a significative influence  on the nature of the scenes \ie~ if significative differences exist between structural features and perceived pleasantness. The significativity is measured with a Student test on two populations;
\item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, linear correlations between these features and $\mathcal{A}_{scene}$ are considered. The Pearson correlation coefficient is used for that purpose, on the basis of the standard methodology used in multidimensional studies;
\item \emph{analysis of the semantic features}: in order to determine if the affective difference between  the i- and ni- scenes has an influence on the scenes composition in terms of sound sources, or if specific classes are more frequently used to simulate a given environment (i or ni), a V-test is considered. Testing is done at every level of abstraction, and separately for texture and event classes. For each class $j$, and each environment type $k$ ($k=\lbrace i,ni\rbrace$), the value $V_{jk}$ is computed as follows:

@@

where $c$ is the number of available classes, $c_k$ the number of classes for a given type of environment $k$, $c_j$ the number of  $j$ classes, and $c_{jk}$ the number of $j$ classes used for an environment $k$. The V-test tests the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significatively different from the ratio $\frac{c_{jk}}{c_k}$. For a given environment $k$, and class $j$, if the hypothesis is rejected, the class $j$ is said to be typical with respect to environment $k$. The typical classes are called \textbf{sound markers};

@@

In this section, relationships that could exist between structural features and perceived pleasantness are investigated. Contrary to the previous section, we do not consider the pleasantness property of a scene on a binary mode (i vs. ni): this is the mean pleasantness $\mathcal{A}_{scene}$ that is considered here as a perceptual feature. The aim is to investigate the level of correlation between structural features and $\mathcal{A}_{scene}$. The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown on Table~\ref{tab:corrStructA}. Relationships between $\mathcal{A}_{scene}$ and the structural features are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and ~\ref{fig:soundlevelf}.

@@

\begin{table}[t]
\centering
\begin{tabular}{l c c c}
               & i-scenes                   & ni-scenes    \\
\hline
$L$            & -0.32 ($p=0.06$)           & \textbf{-0.78} ($p<0.01$)\\
$L(E)$         & -0.20 ($p=0.24$)           & \textbf{-0.75} ($p<0.01$)\\
$L(T)$         & -0.33 ($p=0.05$)           &  -0.00 ($p=0.99$) \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and structural features.}
\label{tab:corrStructA}
\end{table}

>>>>

\newcommand{\gl}[1]{\textcolor{blue}{GL : #1}}
\newcommand{\nm}[1]{\textcolor{magenta}{#1}}
\newcommand{\ml}[1]{\textcolor{red}{#1}}
\newcommand{\jfp}[1]{\textcolor{green}{JFP : #1}}
\newcommand{\ie}{\emph{i.\,e.}}
\newcommand{\Ie}{\emph{I.\,e.}}
\newcommand{\cf}{cf.}
\newcommand{\Cf}{Cf.}
\newcommand{\vs}{\emph{vs.}}
\newcommand{\Vs}{\emph{Vs.}}
\newcommand{\al}{\emph{et~al.}}
\newcommand{\eg}{\emph{e.\,g.}}
\newcommand{\Eg}{\emph{E.\,g.}}
\newcommand{\etc}{\emph{etc.}}
\newcommand{\myfloatalign}{\centering}

@@

    \author{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J-F. Petiot$^{1)}$}

@@

First, a subject can be asked to assess a given sound following a given perceptual scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}. The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}. That being said, those recorded stimuli  can be analysed, what allows the experimenter to gather a coarse
grain physical description of them.

@@

Second, a subject can be asked to describe a given sound environment \cite{guastavino2006ideal, dubois2006cognitive}. A large amount of quantitative and semantic information is then collected about the subject's representation of this type of sound environment. Unfortunately, without any reference to sound data, this representation can hardly be characterized physically.

@@

For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sound sources that compose a sound environment of interest. With such a material, not only the type of sound sources is indeed available for study, but also the exact level and audio waveform for each source together with the structural characteristics of the scene, that is the temporal distribution of the events.

@@

The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}. The experiments  1.a \emph{simulation} and  1.b \emph{pleasantness evaluation} are presented in Section \ref{sec:simulation}. Experiment 2 \emph{pleasantness evaluation after modification of the scenes} is presented in Section \ref{sec:modification}. Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.

@@

\emph{Simscene} is an online digital audio tool whose first version has been developed as part of the HOULE project\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}. It has been designed to run on the popular web browsers \emph{Chrome} and \emph{Firefox}. It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digita audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}. The interface for selecting the sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.

@@

\emph{Simscene} is designed as a simplified audio sequencer, with sequencing parameters specifically chosen for the generation of realistic soundscapes. To do so, the user first selects a sound source using a non verbal selection interface presented in Section~\ref{sec:simscene_ssf}. A track is then created for this sound source within the simulator interface. The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sound sources. Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) to provide some comments about the simulated scene.

@@

The literature usually distinguishes between:
\begin{itemize}
\item {a sound event}: an isolated sound, sparsely distributed whose acoustical characteristics may change with respect to time
\item {a sound texture}: an isolated sound of long duration whose acoustical characteristics are stable with respect to time \cite{saint1995classification}.
\end{itemize}

The simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with the classes of events only and the other with classes of textures only. These two types also have specific scheduling procedures during the simulation process.

@@

Maffiolo \cite{maffiolo_caracterisation_1999} distinguishes two distinct categorization processes, that differentiate themselves with the capacity of the listener to identify sound events. She shows that those processes leads to two abstract cognitive category respectively termed "the event sequences" and "the amorphous sequences".

@@

The event sequences are composed of salient events that can easily be recognized such as \emph{car start} or \emph{male speech}. They arise from a descriptive analysis based on the identification of the sound sources. On the contrary, the amorphous sequences are sound environments where distinct events can hardly be found, such as \emph{traffic hubhub}. They result from a holistic analysis based on global acoustic features.

@@

\ml{fix trim and check why figures goes offline}

\begin{figure}[t]
        \myfloatalign
        \includegraphics[trim={5cm 0 0 0},clip,width=.8\linewidth]{gfx/ch_4/3.png}
       \caption{Hierachical organisation of the isolated sounds used of the simulation.}\label{fig:orgDb}
\end{figure}

@@

The design of this experiment has been validated with a pilot study described in \cite{lafay2014new}.

@@

\begin{itemize}
\item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni- scenes is also present in terms of perceived pleasantness, we look at the existence of differences between $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$ with respect to the two types of scenes. The significativity is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$;
\jfp{ce qu'est Ascene dans ce paragraphe n'est pas clair :
A scene est défini comme la moyenne des scores d'une scène donnée pour tous les participants.

le test, lui, porte sur la moyenne des Ascene pour les deux catégories
même chose pour les sujets}
\item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni- scenes has a significative influence  on the nature of the scenes \ie~ if significative differences exist between structural features and perceived pleasantness. The significativity is measured with a Student test on two populations \jfp{je ne comprends pas bien ce test}
\item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, linear correlations between these features and $\mathcal{A}_{scene}$ are considered. The Pearson correlation coefficient is used for that purpose, on the basis of the standard methodology used in multidimensional studies;
\item \emph{analysis of the semantic features}: in order to determine if the affective difference between  the i- and ni- scenes has an influence on the scenes composition in terms of sound sources, or if specific classes are more frequently used to simulate a given environment (i or ni), a V-test is considered. Testing is done at every level of abstraction, and separately for texture and event classes. For each class $j$, and each environment type $k$ ($k=\lbrace i,ni\rbrace$), the value $V_{jk}$ is computed as follows:

@@

where $c$ is the number of available classes, $c_k$ the number of classes for a given type of environment $k$, $c_j$ the number of  $j$ classes, and $c_{jk}$ the number of $j$ classes used for an environment $k$. \jfp{il faudrait donner une référence pour ce test : bouquin Lebart, Morineau, Piron, STATISTIQUE EXPLORATOIRE MULTIDIMENSIONNELLE
peut être donner les bornes de variations de Vjk pour aider le lecteur à comprendre cet indicateur} \ml{y pas une ref en Anglais ?} The V-test tests the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significatively different from the ratio $\frac{c_{jk}}{c_k}$. For a given environment $k$, and class $j$, if the hypothesis is rejected, the class $j$ is said to be typical with respect to environment $k$. The typical classes are called \textbf{sound markers};\jfp{on ne comprend pas bien l'hypothèse alternative du test.

VOIR BOUQUIN LEBART page 184
ou bien article de Morineau sur le sujet

Vjk est une estimation approchée de la valeur test
}

@@

In this section, relationships that could exist between structural features and perceived pleasantness are investigated. Contrary to the previous section, we do not consider the pleasantness property of a scene on a binary mode (i vs. ni): this is the mean pleasantness $\mathcal{A}_{scene}$ that is considered here as a perceptual feature. The aim is to investigate the level of correlation between structural features and $\mathcal{A}_{scene}$. The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown on Table~\ref{tab:corrStructA}. Relationships between $\mathcal{A}_{scene}$ and the structural features are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and ~\ref{fig:soundlevelf}.  \jfp{en table 2, il devrait y avoir les coefficients de corrélation pour toutes les données, puis pour les deux cas ni et i.
}

@@

\begin{table}[t]
\centering
\begin{tabular}{l c c c}
               & i-scenes                   & ni-scenes    \\
\hline
$L$            & -0.32 ($p=0.06$)           & \textbf{-0.78} ($p<0.01$)\\
$L(E)$         & -0.20 ($p=0.24$)           & \textbf{-0.75} ($p<0.01$)\\
$L(T)$         & -0.33 ($p=0.05$)           &  -0.00 ($p=0.99$) \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and structural features.\jfp{rajouter une colonne corélation globale}}
\label{tab:corrStructA}
\end{table}
